Hyperplane Elimination for Quickly Enumerating Local Optima

### 2016-01-08 Hyperplane Enumeration/main.tex
A few more

\begin{abstract}
Examining the properties of local optima is a common method for understanding
combinatorial-problem landscapes.
Unfortunately,
exhaustive algorithms for finding local optima are limited to very small problem sizes.
We propose a method for exploiting problem structure
to skip hyperplanes
that cannot contain local optima, allowing runtime to scale with
the number of local optima instead of with the landscape size.
We prove optimality for linear problems and the Concatenated Traps, and we provide
empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
We further refine this method to find solutions
that cannot be improved by flipping $r$ or fewer bits, which counterintuitively
can reduce total runtime. While previous methods were limited to
landscapes with at most $2^{34}$ binary strings, hyperplane elimination can enumerate the same problems with
$2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
%We believe this increase in size will help further our understanding of complex real-world landscapes.

@@

\subsection{$k$-Bound Separable Problems}
A $k$-bound separable problem is any problem that is composed of non-overlapping
subfunctions, each using using $k$ or fewer bits. Formally this means
$\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.

@@

Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
by finding all local optima in larger and larger hyperplanes. Initially, $f_0$
is enumerated to find all local optima in the hyperplane where all $f_i, i>0$ remain
fixed. This process ends when a carry operation modifies a bit of $f_1$. At
this point, $f_1$ is enumerated such that each time $f_1$ contains no improving moves
$f_0$ is enumerated again. This finds all local optima in the hyperplane where $f_i, i>1$ remain
fixed. We represent the time required to find all local optima in the hyperplane
with $f_j, j>i$ fixed as $T(i)$. $T(0)=2^k$ as all values of $f_0$ must be enumerated, and
$T(i) = |l_i|*T(i-1)+2^k$.
Each setting in $l_i$ exposes a new hyperplane
that can contain local optima, which causes the recursive call to $T(i-1)$.
%As all $|s_i| \leq k$ and $k$ is a
%constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
The time required to find all local optima is therefore
$T(M-1) = 2^k\sum_{j=0}^{M-1} \prod_{i=M-j}^{M-1} |l_i|$.
\end{proof}

@@

%TODO Reviewer 1: State programming language and statistical tools
Each algorithm was given a maximum of four hours to find all $r$-bit local optima.
Runs were distributed across a cluster using 2.5GHz Intel Xeon E5-2670v2 processors.
Any run which failed was rerun once to prevent cluster issues from introducing
noise. For the same reason, any run which was over ten times slower than the next slowest run of the
same configuration with a different seed was rerun. This resulted in 33 out of 36,000 runs being rerun.
If the algorithm failed to complete in the alloted time for an instance,
we declared it unsuccessful for that combination of $r$ and $N$ of that problem. All results
we report are for successful combinations. For timing purposes, local optima were counted but not
recorded. All of our code, data, and statistical analysis can be downloaded from our
website.\footnote{\url{https://name.of.site/removed/to/be/anonymous/}}
%TODO Make this the actual web page

>>>>

\begin{abstract}
Examining the properties of local optima is a common method for understanding
combinatorial-problem landscapes.
Unfortunately,
exhaustive algorithms for finding local optima are limited to very small problem sizes.
We propose a method for exploiting problem structure
to skip hyperplanes
that cannot contain local optima, allowing runtime to scale with
the number of local optima instead of with the landscape size.
We prove optimality for linear problems and Concatenated Traps, and we provide
empirical evidence of optimality on NKq Landscapes and Ising Spin Glasses.
We further refine this method to find solutions
that cannot be improved by flipping $r$ or fewer bits, which counterintuitively
can reduce total runtime. While previous methods were limited to
landscapes with at most $2^{34}$ binary strings, hyperplane elimination can enumerate the same problems with
$2^{77}$ binary strings, and find all 4-bit local optima to problems with $2^{200}$ binary strings.
%We believe this increase in size will help further our understanding of complex real-world landscapes.

@@

\subsection{$k$-Bound Separable Problems}
A $k$-bound separable problem is any problem that is composed of non-overlapping
subfunctions, each using using $k$ or fewer bits. Formally this means
$\forall_i |s_i| \leq k$ and $\forall_{i \neq j} s_i \cap s_j = \emptyset$.
Let $l_i$ be the set of ways $f_i$ can be set
such that it contains no fitness improving moves.

@@

Once the first local optimum is found, Algorithm~\ref{alg-enumerate} proceeds
by finding all local optima in larger and larger hyperplanes. Initially, $f_0$
is enumerated to find all local optima in the hyperplane where all $f_i, i>0$ remain
fixed. This process ends when a carry operation modifies a bit of $f_1$. At
this point, $f_1$ is enumerated such that each time $f_1$ contains no improving moves
$f_0$ is enumerated again. This finds all local optima in the hyperplane where $f_i, i>1$ remain
fixed. $T(i)$ represents the time required to find all local optima in the hyperplane
with $f_j, j>i$ fixed. $T(0)=2^k$ as all values of $f_0$ must be enumerated, and
$T(i) = |l_i|*T(i-1)+2^k$.
Each setting in $l_i$ exposes a new hyperplane
that can contain local optima, which causes the recursive call to $T(i-1)$.
%As all $|s_i| \leq k$ and $k$ is a
%constant, $\forall_i |l_i| \leq c \leq 2^k$ meaning that $c$ is a constant.
The time required to find all local optima is therefore
$T(M-1) = 2^k\sum_{j=0}^{M-1} \prod_{i=M-j}^{M-1} |l_i|$.
\end{proof}

@@

%TODO Reviewer 1: State programming language and statistical tools
Each algorithm was given a maximum of four hours to find all $r$-bit local optima.
Runs were distributed across a cluster using 2.5GHz Intel Xeon E5-2670v2 processors.
Any run which failed was rerun once to prevent cluster issues from introducing
noise. For the same reason, any run which was over ten times slower than the next slowest run of the
same configuration with a different seed was rerun. This resulted in 33 out of 36,000 runs being rerun.
If the algorithm failed to complete in the alloted time for an instance,
we declared it unsuccessful for that combination of $r$ and $N$ of that problem. All results
we report are for successful combinations. For timing purposes, local optima were counted but not
recorded. All of our code, data, and statistical analysis can be downloaded from our
website.\footnote{\url{https://github.com/brianwgoldman/Enumerate-Local-Optima/releases}}
%TODO Make this the actual web page
